---
title: "Gender Pay Gap in Australia: Industry-Level Trends and Analysis"
format: pdf
  toc: true
editor: visual
editor_options: 
  chunk_output_type: console
---
<<<<<<< HEAD

<!-- START: Executive Summary -->

## Executive Summary

<!-- END: Executive Summary -->

<!-- START: Introduction -->

## Introduction

<!-- END: Introduction-->

<!-- START: Methodology -->

## Methodology
```{r}
#| echo: false

library(fpp3)

# Data cleaning and preperation

# Gender wage gap data
gender_wage_gap <- read.csv("Data/gender wage gap across time.csv") %>% 
  select(Reference.area, OBS_VALUE, TIME_PERIOD) %>% 
  rename(Country = Reference.area, Gender_Wage_Gap = OBS_VALUE, Year = TIME_PERIOD) %>% 
  arrange(Year)

gender_wage_gap <- gender_wage_gap %>%
  group_by(Country, Year) %>%
  summarise(Gender_Wage_Gap = mean(Gender_Wage_Gap), .groups = "drop")

gender_wage_gap <- gender_wage_gap %>%
  mutate(Country = ifelse(Country == "European Union (27 countries)", "European Union", Country))

# GDP per capita data
gdp <- read.csv("Data/gdp per capita.csv") %>% 
  select(-X.1) %>% 
  rename(Country = X)

gdp <- gdp %>%
  pivot_longer(
    cols = starts_with("X"),
    names_to = "Year",
    values_to = "GDP_per_capita"
  ) %>%
  mutate(year = as.integer(sub("X", "", Year))) 

gdp <- gdp %>% 
  select(Country, GDP_per_capita, year) %>% 
  rename(Year = year)

# Merge the two datasets
data <- left_join(gender_wage_gap, gdp, by = c("Country", "Year"))

# Gender stats data
gender <- read.csv("Data/gender stats.csv") 

gender <- gender[-c(15,16,17,18,19), ]

gender <- gender %>% 
  select(-Country.Code, -Series.Code) %>% 
  rename(y = Series.Name, Country = Country.Name)

# Labor force participation, female (% of female pop aged 15-64)
labor_force_participation <- gender %>% 
  filter(y == "Labor force participation rate, female (% of female population ages 15-64) (modeled ILO estimate)") %>% 
  select(-y) %>% 
  pivot_longer(
    cols = -Country,
    names_to = "Year",
    values_to = "Labour_force_participation"
  ) %>%
  mutate(Year = as.integer(sub(".*?(\\d{4}).*", "\\1", Year))) 

# Part time employment, female (% of total female employment)
part_time_employment <- gender %>% 
  filter(y == "Part time employment, female (% of total female employment)") %>% 
  select(-y) %>% 
  pivot_longer(
    cols = -Country,
    names_to = "Year",
    values_to = "Part_time_employment"
  ) %>%
  mutate(Year = as.integer(sub(".*?(\\d{4}).*", "\\1", Year)))
  
# Merge datasets
data <- left_join(data, labor_force_participation, by = c("Country", "Year"))

data <- left_join(data, part_time_employment, by = c("Country", "Year")) 

```












<!-- END: Methodology -->

<!-- START: Results -->

## Results

How does the gender wage gap trend differ across high-income vs. low-income countries?

```{r}
# Clean GDP_per_capita: remove commas and convert to numeric
data <- data %>%
  mutate(GDP_per_capita = as.numeric(gsub(",", "", GDP_per_capita)))

# Now create the Income_Group variable
data <- data %>%
  group_by(Year) %>%
  mutate(Income_Group = if_else(GDP_per_capita >= median(GDP_per_capita, na.rm = TRUE),
                                "High Income", "Low Income")) %>%
  ungroup()


# Plot gender wage gap trends by income group
data %>%
  group_by(Year, Income_Group) %>%
  summarise(Avg_Gender_Gap = mean(Gender_Wage_Gap, na.rm = TRUE)) %>%
  ggplot(aes(x = Year, y = Avg_Gender_Gap, color = Income_Group)) +
  geom_line(size = 1.2) +
  labs(title = "Gender Wage Gap Trends by Income Group",
       y = "Average Gender Wage Gap (%)",
       x = "Year") +
  theme_minimal()

```

How have gender wage gaps changed globally from 2000 to 2020?

```{r}
# Global average trend
data %>%
  group_by(Year) %>%
  summarise(Global_Gender_Gap = mean(Gender_Wage_Gap, na.rm = TRUE)) %>%
  ggplot(aes(x = Year, y = Global_Gender_Gap)) +
  geom_line(color = "darkblue", size = 1.2) +
  labs(title = "Global Gender Wage Gap Trend (2000–2020)",
       y = "Global Average Gender Wage Gap (%)",
       x = "Year") +
  theme_minimal()

```


Are countries with higher labour force participation rates more likely to have smaller gender wage gaps?

```{r}
# Scatter plot with linear trend line
data %>%
  ggplot(aes(x = Labour_force_participation, y = Gender_Wage_Gap)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  labs(title = "Relationship Between Labour Force Participation and Gender Wage Gap",
       x = "Labour Force Participation (%)",
       y = "Gender Wage Gap (%)") +
  theme_minimal()

```

The analysis reveals a clear global decline in the gender wage gap from 2000 to 2020, dropping from over 23% to below 17%. This suggests global progress toward pay equality. When comparing countries by income group, high-income countries consistently have lower gender wage gaps than low-income ones. Both groups show improvement over time, but the gap remains larger in low-income countries, indicating that economic development may support gender wage equality.

The scatterplot shows a negative relationship between labour force participation and the gender wage gap. Countries with higher labour force participation tend to have smaller wage gaps, possibly due to increased female presence in skilled roles and more equitable workplace policies. However, some variation exists, suggesting other contributing factors.

Overall, while progress is evident, wage disparities persist—especially in lower-income countries and those with lower labour participation.


<!-- END: Results -->
