---
title: "What Explains the Gender Pay Gap Across Countries?"
author:
  - name: Muhammad Nor Hakim bin Mad Yusof (28693310)
  - name: Chanya Kitprapa (34684824)
  - name: Janhavi Bayanwar (35579331)
format:
  pdf:
    toc: true
    toc-depth: 2          # Includes ## subheadings (up to level 2)
    number-sections: true
    fig-cap-location: bottom
    tbl-cap-location: top
    keep-tex: true        # Helpful for debugging LaTeX issues
    documentclass: article
    fontsize: 11pt
    linestretch: 1.5
    geometry: margin=1in
    colorlinks: true
    mainfont: "Times New Roman"
    monofont: "Courier New"
    bibliography: references.bib
editor: visual
nocite: |
  @ourworldindata_gender_inequality
  @oecd_gender_equality_work
  @worldbank_gender_statistics
execute:
  echo: false
  warning: false
  message: false
---

# Executive Summary

This report investigates the gender wage gap across seven OECD countries, analyzing how economic indicators and labor force characteristics shape earnings disparities between men and women. Despite rising female education levels and labor force participation, the wage gap remains persistent—particularly in countries like Japan and South Korea—highlighting the influence of structural and policy-driven factors. Nations such as Sweden and the European Union demonstrate that progressive labor policies can contribute to more equitable outcomes. The findings suggest that addressing occupational segregation, care responsibilities, and wage transparency is essential for closing the gender pay gap.

# Introduction

The gender pay gap, a measure of the difference in median earnings between men and women as a percentage of men's earnings, is a widely used indicator of economic inequality. It captures disparities not only in pay for equal work, but also in broader structural dynamics such as occupational segregation, unpaid care burdens, and labor force participation. Importantly, a visible pay gap does not always reflect direct discrimination, nor does its absence guarantee equality. Discrimination may still influence career choices or discourage women from entering or staying in the workforce altogether. Conversely, large pay gaps can exist even in societies with formal equality if women are concentrated in lower-paying roles or sectors.

This complexity makes the gender pay gap a useful but incomplete signal—it reflects inequality, but does not explain it. Across countries, the size and persistence of the gap vary widely, raising the question of what social, cultural, and institutional factors are driving these differences. While education levels between men and women have converged globally, disparities in earnings remain. This suggests that deeper forces—such as caregiving norms, workplace structures, and policy frameworks—play a crucial role. This report explores how and why the gender pay gap differs across countries, identifying key indicators that help explain cross-national patterns of inequality.

# Methodology

```{r}
# Load packages

library(tidyverse)
library(fpp3)
library(kableExtra)
library(dplyr)
library(stringr)

# Data cleaning and preparation
# Gender wage gap data

gender_wage_gap <- read.csv("gender wage gap across time.csv") %>%
  select(Reference.area, OBS_VALUE, TIME_PERIOD) %>% 
  rename(Country = Reference.area, Gender_Wage_Gap = OBS_VALUE, Year = TIME_PERIOD) %>% 
  arrange(Year)

gender_wage_gap <- gender_wage_gap %>%
  group_by(Country, Year) %>%
  summarise(Gender_Wage_Gap = mean(Gender_Wage_Gap), .groups = "drop")

gender_wage_gap <- gender_wage_gap %>%
  mutate(Country = ifelse(Country == "European Union (27 countries)", "European Union", Country))

# GDP per capita data
gdp <- read.csv("gdp per capita.csv") %>%
  select(-X.1) %>% 
  rename(Country = X)

gdp <- gdp %>%
  pivot_longer(
    cols = starts_with("X"),
    names_to = "Year",
    values_to = "GDP_per_capita"
  ) %>%
  mutate(year = as.integer(sub("X", "", Year))) 

gdp <- gdp %>% 
  select(Country, GDP_per_capita, year) %>% 
  rename(Year = year)

# Merge the two datasets
data <- left_join(gender_wage_gap, gdp, by = c("Country", "Year"))

# Gender stats data

gender <- read.csv("gender stats.csv") 


gender <- gender[-c(15,16,17,18,19), ]

gender <- gender %>% 
  select(-Country.Code, -Series.Code) %>% 
  rename(y = Series.Name, Country = Country.Name)

# Labor force participation, female (% of female pop aged 15-64)
labor_force_participation <- gender %>% 
  filter(y == "Labor force participation rate, female (% of female population ages 15-64) (modeled ILO estimate)") %>% 
  select(-y) %>% 
  pivot_longer(
    cols = -Country,
    names_to = "Year",
    values_to = "Labour_force_participation"
  ) %>%
  mutate(Year = as.integer(sub(".*?(\\d{4}).*", "\\1", Year))) 

# Part time employment, female (% of total female employment)
part_time_employment <- gender %>% 
  filter(y == "Part time employment, female (% of total female employment)") %>% 
  select(-y) %>% 
  pivot_longer(
    cols = -Country,
    names_to = "Year",
    values_to = "Part_time_employment"
  ) %>%
  mutate(Year = as.integer(sub(".*?(\\d{4}).*", "\\1", Year)))
  
# Merge datasets
data <- left_join(data, labor_force_participation, by = c("Country", "Year"))

data <- left_join(data, part_time_employment, by = c("Country", "Year")) 

# Educational attainment, at least Bachelor's or equivalent, population 25+, female (%) (cumulative)
female_bachelors <- read.csv("female bachelors attainment.csv") %>% 
  rename(Country = X) %>% 
  select(-X.1) %>% 
  mutate(Country = ifelse(Country == "Korea, Rep.", "Korea", Country)) %>% 
  pivot_longer(
    cols = starts_with("X"),
    names_to = "Year",
    values_to = "Female_bachelors") %>%
  mutate(Year = as.integer(sub("X", "", Year))) 

# Merge datasets
data <- left_join(data, female_bachelors, by = c("Country", "Year"))

# Convert data str
data <- data %>% 
  mutate(Labour_force_participation = as.numeric(Labour_force_participation))
```
This study investigates cross-country differences in the gender wage gap by analyzing trends from 2000 to 2020 across seven countries and regions: Australia, United Kingdom (U.K), United States (U.S), Sweden, Japan, South Korea, and the European Union (EU). The gender wage gap is measured as the difference in median earnings between men and women, expressed as a percentage of male earnings. This metric captures broader structural inequality beyond direct wage discrimination, and serves as the dependent variable in our analysis.

To explore possible explanatory factors, we compiled four key indicators for each country and year:

-   GDP per capita (US$, 2015)
-   Female labor force participation rate (% of female population aged 15-64)
-   Part-time employment (% of total female employment)
-   Educational attainment of females aged 25+ at bachelor's level or equivalent (% of females).

All variables were sourced from the World Bank and Organisation for Economic Co-operation and Development (OECD) gender gata portals and merged into a panel dataset by country and year. Some variables for certain countries in certain years are missing due to incomplete data collection by these bodies.

Time series plots were used to observe trends and variation and to also compare the gender wage gap to the 2023 OECD average of 11% - our threshold. All data wrangling and visualization were conducted in R using the tidyverse and ggplot2 packages.

## Snapshot of dataset

```{r}
#| label: tbl-snapshot
#| tbl-cap: "Summary of final dataset used for analysis"

data %>% 
  filter(Year == 2020) %>% 
  kable()
```
A snapshot of the dataset, including 2020 values for each country, is presented in @tbl-snapshot.

## Gender wage gap

```{r}
#| label: fig-medianwagegap
#| fig-cap: "Gender wage gap by country (2000-2020)"
#| fig-align: "center"
#| fig-pos: "H"

ggplot(data, aes(x = Year, y = Gender_Wage_Gap, color = Country)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  labs(
    x = "Year",
    y = "Gender Wage Gap (%)",
    color = "Country"
      ) +
  theme_minimal() +
  geom_hline(yintercept = 11, linetype = "dashed", color = "red") +
  annotate("text", x = Inf, y = 11, label = "OECD average", 
           hjust = 1.1, vjust = -0.5, color = "red", size = 3.5)
```
@fig-medianwagegap shows that while the gender wage gap declined across all seven countries from 2000 to 2020, the rate and extent of improvement varied significantly. Korea and Japan maintained the highest wage gaps throughout the period, despite gradual declines, whereas Sweden and the European Union exhibited the lowest and most consistent reductions. The U.K, U.S, and Australia showed moderate progress, with notable year-to-year fluctuations.

# Results

## Evolution of women's education attainment across countries (2013-2020)

```{r}
#| label: fig-femaleeducation
#| fig-cap: "Evolution of female education attainment across countries (2013-2020)"
#| fig-align: "center"
#| fig-pos: "H"

education_filtered <- data %>%
  filter(Year >= 2013, !Country %in% c("European Union", "Japan", "Korea"))

education_filtered <- education_filtered %>% 
  mutate(Female_bachelors = as.numeric(Female_bachelors))

ggplot(education_filtered, aes(x = Year, y = Female_bachelors, color = Country, group = Country)) +
  geom_line(size = 1.2) +
  labs(
    x = "Year",
    y = "Female Bachelor's or Equivalent Attainment (%)",
    color = "Country"
      ) +
  theme_minimal()
```
From 2013 to 2020, female education attainment rose across all countries, surpassing 36% by 2020 (@fig-femaleeducation). Yet, the persistent gender wage gap suggests that education alone cannot close pay disparities without tackling deeper structural and occupational inequalities.

## Gender wage gap VS. GDP per capita (US$, 2015)

```{r}
#| label: fig-wagegapvsgdp
#| fig-cap: "Gender wage gap vs GDP per capita across countries (2000-2020)"
#| fig-align: "center"
#| fig-pos: "H"

data_filtered <- data %>% 
  filter(Country !="Korea")

data_gdp <- data_filtered %>%
  mutate(GDP_per_capita = str_replace_all(GDP_per_capita, "[^0-9.]", ""),
         GDP_per_capita = as.numeric(GDP_per_capita))

ggplot(data_gdp, aes(x = GDP_per_capita, y = Gender_Wage_Gap, colour = Country)) +
  geom_point(size = 4) +
  labs(
    x = "GDP per Capita (US$, 2015)",
    y = "Gender Wage Gap (%)",
    color = "Country"
  ) +
  theme_minimal() +
  geom_hline(yintercept = 11, linetype = "dashed", color = "red") +
  annotate("text", x = Inf, y = 11, label = "OECD average", 
           hjust = 1.1, vjust = -0.5, color = "red", size = 3.5)
```
In @fig-wagegapvsgdp, the gender wage gap tends to decrease with higher GDP per capita, with Sweden and the EU falling below the OECD average. However, persistently high gaps in countries like Japan and the U.S suggest that income alone does not explain wage inequality.

## Gender wage gap VS. Female labor force participation

```{r}
#| label: fig-wagegapvsflabpart
#| fig-cap: "Gender wage gap vs Female labor force participation (2000-2020)"
#| fig-align: "center"
#| fig-pos: "H"

data_filtered <- data %>% 
  filter(Country !="Korea")
  
ggplot(data_filtered, aes(x = Labour_force_participation, y = Gender_Wage_Gap, color = Year)) +
  geom_point(size = 2, alpha = 0.7) +
  facet_wrap(~ Country) +
  scale_color_viridis_c() +
  labs(
    x = "Female Labor Force Participation (%)",
    y = "Gender Wage Gap (%)",
    color = "Year"
      ) +
  theme_minimal() +
  scale_x_continuous(breaks = seq(2000, 2020, by = 5)) +
  geom_hline(yintercept = 11, linetype = "dashed", color = "red") +
  annotate("text", x = Inf, y = 11, label = "OECD average", 
           hjust = 1.1, vjust = -0.5, color = "red", size = 3.5)
```
@fig-wagegapvsflabpart shows that in most countries, increases in female labor force participation over time are associated with declines in the gender wage gap. However, the strength and consistency of this relationship vary: countries like the U.K and EU show a strong negative association, while the U.S displays a relatively flat or even slightly positive trend.

## Gender wage gap VS. Part-time female employment

```{r}
#| label: fig-wagegapvsparttime
#| fig-cap: "Gender wage gap vs Part-time female employment (2000-2020)"
#| fig-align: "center"
#| fig-pos: "H"

data_filtered_new <- data %>% 
  filter(!Country %in% c("Korea", "Japan"))
  
ggplot(data_filtered, aes(x = Labour_force_participation, y = Gender_Wage_Gap, color = Year)) +
  geom_point(size = 2, alpha = 0.7) +
  facet_wrap(~ Country) +
  scale_color_viridis_c() +
  labs(
    x = "Part-time female employment (%)",
    y = "Gender Wage Gap (%)",
    color = "Year"
      ) +
  theme_minimal() +
  scale_x_discrete(breaks = seq(2000, 2020, by = 5)) +
  geom_hline(yintercept = 11, linetype = "dashed", color = "red") +
  annotate("text", x = Inf, y = 11, label = "OECD average", 
           hjust = 1.1, vjust = -0.5, color = "red", size = 3.5)
```
@fig-wagegapvsparttime shows a positive association between female part-time employment and the gender wage gap in several countries, especially Japan and the U.K. However, Sweden demonstrates that even with relatively high part-time rates, strong labor protections may have helped sustain a lower wage gap over time.

# Discussion

The gender wage gap remains a persistent global issue, with substantial variation across countries and over time. This report analyzed the gender wage gap across several developed OECD nations to assess the extent of income inequality. These countries were chosen to provide a somewhat even basis of analysis due to their economic maturity, whilst having a nuanced comparison due to their differing industrial and policy make ups. 

Countries such as South Korea and Japan consistently exhibit some of the highest wage gaps—often exceeding 20%—reflecting entrenched structural inequalities and traditional gender norms in the workplace. In contrast, nations like Sweden and the European Union demonstrate significantly lower gaps, supported by progressive labor policies, subsidized childcare, and equitable parental leave. Although the overall trend points to a gradual decline in the wage gap, the pace of progress has slowed in recent years. Some countries exhibit fluctuating trajectories rather than steady improvement, suggesting that changes in policy, female labor force participation, and broader economic conditions all influence wage equity. Notably, the gender wage gap metric does not account for variations in occupation, industry, seniority, or hours worked—factors that can obscure or amplify underlying discrimination—highlighting the complexity of interpreting this indicator in isolation.

# Conclusion

This report has examined cross-national differences in the gender wage gap, highlighting how structural, economic, and institutional factors contribute to persistent earnings disparities between men and women. While countries such as Sweden and the European Union have achieved relatively low wage gaps, others—including Japan and South Korea—continue to report high levels of inequality despite economic growth. Although female education attainment has improved and labor force participation has increased, these advances alone have not closed the wage gap, indicating that deeper structural reforms are needed. The analysis underscores that the gender wage gap is not solely a function of individual qualifications or productivity, but also of entrenched occupational segregation, unequal caregiving responsibilities, and insufficient policy support.

# Recommendations

Based on our analysis, these are the suggested policy actions that may improve the narrowing of the gender wage gap:

- Implement Gender-Responsive Labor Policies

  Governments should strengthen laws that promote equal pay, flexible work arrangements, and inclusive parental leave to support women’s full economic participation.

- Invest in Childcare and Family Support Services

  Expanding access to affordable childcare and shared parental leave can reduce the disproportionate burden of unpaid care work on women and facilitate career continuity.

- Increase Transparency and Data Reporting

  Mandating the disclosure of wage data by gender and occupation can help identify disparities within organizations and encourage accountability.

- Promote Female Representation in Leadership

  Policies that support the advancement of women into senior and decision-making roles can help disrupt structural barriers and narrow wage gaps over time.

- Complement Wage Gap Metrics with Contextual Indicators

  Future analyses should integrate additional variables—such as hours worked, occupation type, and industry segmentation—to better understand the underlying causes of pay disparities and   guide targeted interventions.
  
# References
