---
title: "Gender Pay Gap in Australia: Industry-Level Trends and Analysis"
format:
  html:
    toc: true
  pdf:
    toc: true
editor: visual
editor_options: 
  chunk_output_type: console
---

<!-- START: Executive Summary -->
## Executive Summary




<!-- END: Executive Summary -->
<!-- START: Introduction -->
## Introduction

<!-- END: Introduction-->
<!-- START: Methodology -->
## Methodology
```{r}
#| echo: false

library(fpp3)
library(kableExtra)
library(dplyr)
library(stringr)

# Data cleaning and preperation

# Gender wage gap data

gender_wage_gap <- read.csv("Data/gender wage gap across time.csv") %>%
  select(Reference.area, OBS_VALUE, TIME_PERIOD) %>% 
  rename(Country = Reference.area, Gender_Wage_Gap = OBS_VALUE, Year = TIME_PERIOD) %>% 
  arrange(Year)

gender_wage_gap <- gender_wage_gap %>%
  group_by(Country, Year) %>%
  summarise(Gender_Wage_Gap = mean(Gender_Wage_Gap), .groups = "drop")

gender_wage_gap <- gender_wage_gap %>%
  mutate(Country = ifelse(Country == "European Union (27 countries)", "European Union", Country))

# GDP per capita data
gdp <- read.csv("Data/gdp per capita.csv") %>%
  select(-X.1) %>% 
  rename(Country = X)

gdp <- gdp %>%
  pivot_longer(
    cols = starts_with("X"),
    names_to = "Year",
    values_to = "GDP_per_capita"
  ) %>%
  mutate(year = as.integer(sub("X", "", Year))) 

gdp <- gdp %>% 
  select(Country, GDP_per_capita, year) %>% 
  rename(Year = year)

# Merge the two datasets
data <- left_join(gender_wage_gap, gdp, by = c("Country", "Year"))

# Gender stats data

gender <- read.csv("Data/gender stats.csv") 


gender <- gender[-c(15,16,17,18,19), ]

gender <- gender %>% 
  select(-Country.Code, -Series.Code) %>% 
  rename(y = Series.Name, Country = Country.Name)

# Labor force participation, female (% of female pop aged 15-64)
labor_force_participation <- gender %>% 
  filter(y == "Labor force participation rate, female (% of female population ages 15-64) (modeled ILO estimate)") %>% 
  select(-y) %>% 
  pivot_longer(
    cols = -Country,
    names_to = "Year",
    values_to = "Labour_force_participation"
  ) %>%
  mutate(Year = as.integer(sub(".*?(\\d{4}).*", "\\1", Year))) 

# Part time employment, female (% of total female employment)
part_time_employment <- gender %>% 
  filter(y == "Part time employment, female (% of total female employment)") %>% 
  select(-y) %>% 
  pivot_longer(
    cols = -Country,
    names_to = "Year",
    values_to = "Part_time_employment"
  ) %>%
  mutate(Year = as.integer(sub(".*?(\\d{4}).*", "\\1", Year)))
  
# Merge datasets
data <- left_join(data, labor_force_participation, by = c("Country", "Year"))

data <- left_join(data, part_time_employment, by = c("Country", "Year")) 


```

# Educational attainment, at least Bachelor's or equivalent, population 25+, female (%) (cumulative)

```{r}

female_bachelors <- read.csv("Data/female bachelors attainment.csv") %>% 
  rename(Country = X) %>% 
  select(-X.1) %>% 
  mutate(Country = ifelse(Country == "Korea, Rep.", "Korea", Country)) %>% 
  pivot_longer(
    cols = starts_with("X"),
    names_to = "Year",
    values_to = "Female_bachelors") %>%
  mutate(Year = as.integer(sub("X", "", Year))) 

# Merge datasets
data <- left_join(data, female_bachelors, by = c("Country", "Year"))

# Convert data str
data <- data %>% 
  mutate(Labour_force_participation = as.numeric(Labour_force_participation))


```


This study investigates cross-country differences in the gender wage gap by analyzing trends from 2000 to 2020 across seven countries: Australia, United Kingdom, United States, Sweden, Japan, South Korea, and the European Union. The gender wage gap is measured as the difference in median earnings between men and women, expressed as a percentage of male earnings. This metric captures broader structural inequality beyond direct wage discrimination, and serves as the dependent variable in our analysis.

To explore possible explanatory factors, we compiled four key indicators for each country and year: 

- GDP per capita (2015 USD) 
- Female labor force participation rate (ages 15–64)
- Part-time employment as a percentage of total female employment
- Female educational attainment at or above the bachelor's level (ages 25+). 

All variables were sourced from the World Bank and OECD Gender Data Portals and merged into a panel dataset by country and year. Some variables for certain countries in certain years are missing due to incomplete data collection by these bodies.

Descriptive statistics and time series plots were used to observe trends and variation. All data wrangling and visualization were conducted in R using the tidyverse and ggplot2 packages. Figures and tables are embedded in this Quarto document and referenced dynamically using Quarto cross-referencing syntax.

```{r}
#| label: tbl-summary
#| tbl-cap: "Summary of final dataset used for analysis"

data %>% 
  filter(Year == 2020) %>% 
  kable()
```
A summary of the dataset, including 2020 values for each country, is presented in @tbl-summary.

```{r}
#| label: fig-medianwagegap
#| fig-cap: "Median gender wage gap by country (2000-2020)"

ggplot(data, aes(x = Year, y = Gender_Wage_Gap, color = Country)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  labs(
    title = "Gender Wage Gap",
    x = "Year",
    y = "Median Gender Wage Gap (%)",
    color = "Country"
  ) +
  theme_minimal()
```
@fig-medianwagegap shows that while the median gender wage gap declined across all seven countries from 2000 to 2020, the rate and extent of improvement varied significantly. Korea and Japan maintained the highest wage gaps throughout the period, despite gradual declines, whereas Sweden and the European Union exhibited the lowest and most consistent reductions. The United Kingdom, United States, and Australia showed moderate progress, with notable year-to-year fluctuations.


```{r}
#| label: fig-wagegapvsgdp
#| fig-cap: "Median gender wage gap vs GDP per capita across countries (2000-2020)"

data_filtered <- data %>% 
  filter(Country !="Korea")

data_gdp <- data_filtered %>%
  mutate(GDP_per_capita = str_replace_all(GDP_per_capita, "[^0-9.]", ""),
         GDP_per_capita = as.numeric(GDP_per_capita))

ggplot(data_gdp, aes(x = GDP_per_capita, y = Gender_Wage_Gap, colour = Country)) +
  geom_point(size = 4) +
  labs(
    x = "GDP per Capita (2015, $US)",
    y = "Median Gender Wage Gap (%)",
    color = "Country"
  ) +
  theme_minimal()
```
@fig-wagegapvsgdp shows that GDP per capita does not have a consistent or linear relationship with the median gender wage gap across countries. While higher-income countries like Sweden and the European Union tend to exhibit lower wage gaps, the United States and Australia maintain relatively high wage gaps despite high GDP per capita, indicating that economic prosperity alone does not ensure gender pay equity.

```{r}
#| label: fig-wagegapvsflabpart
#| fig-cap: "Median gender wage gap vs Female labor force participation (2000-2020)"

data_filtered <- data %>% 
  filter(Country !="Korea")
  
ggplot(data_filtered, aes(x = Labour_force_participation, y = Gender_Wage_Gap, color = Year)) +
  geom_point(size = 2, alpha = 0.7) +
  facet_wrap(~ Country) +
  scale_color_viridis_c() +
  labs(
    title = "Evolution of Gender Wage Gap vs. Labor Participation (2000–2020)",
    x = "Female Labor Force Participation (%)",
    y = "Median Gender Wage Gap (%)",
    color = "Year"
  ) +
  theme_minimal() +
  scale_x_continuous(breaks = seq(2000, 2020, by = 5))
```
@fig-wagegapvsflabpart shows that in most countries, increases in female labor force participation over time are associated with declines in the median gender wage gap. However, the strength and consistency of this relationship vary: countries like the UK and EU show a strong negative association, while the US displays a relatively flat or even slightly positive trend.

<!-- END: Methodology -->
<!-- START: Results -->
## Results

Across the observed countries, the median gender wage gap declined between 2000 and 2020, though the rate and consistency of improvement varied. South Korea and Japan maintained relatively high wage gaps throughout the period, while Sweden and the European Union recorded the lowest and most stable reductions. Australia, the UK, and the US showed moderate but uneven progress. This variation is illustrated in fig-wagegapvsgd which compares the gender wage gap to GDP per capita. The figure highlights that higher national income does not necessarily translate to greater gender pay equity. For instance, the United States and Australia report high GDP per capita but continue to exhibit significant wage disparities. Conversely, Sweden and the EU combine strong economic performance with narrower wage gaps. These findings suggest that while economic development may influence gender pay outcomes, it is not the sole determining factor. Other structural elements, such as labor market policies, social norms, and institutional support for gender equality, are likely to shape the extent of wage disparities across countries.

```{r}
#| label: fig-wagegapvsgdp
#| fig-cap: "Median gender wage gap vs GDP per capita across countries (2000-2020)"

data_filtered <- data %>% 
  filter(Country !="Korea")

data_gdp <- data_filtered %>%
  mutate(GDP_per_capita = str_replace_all(GDP_per_capita, "[^0-9.]", ""),
         GDP_per_capita = as.numeric(GDP_per_capita))

ggplot(data_gdp, aes(x = GDP_per_capita, y = Gender_Wage_Gap, colour = Country)) +
  geom_point(size = 4) +
  labs(
    x = "GDP per Capita (2015, $US)",
    y = "Median Gender Wage Gap (%)",
    color = "Country"
  ) +
  theme_minimal()
```


<!-- END: Results -->

<!-- START: Discussion -->
## Discussion

The gender wage gap which varies greatly between nations and over time, is still a persistent global problem. The report analyzes the percentage difference in the earnings of females compared to males to assess salary discrepancies across a large sample of OECD nations. Due to ingrained structural disparities and gender norms in the workplace, countries like South Korea and Japan continue to have some of the highest gaps, frequently surpassing 20%.
In contrast nations like Sweden and the European Union on the other hand have significantly lower wage gaps, owing to progressive labour laws, subsidized child care and equal parental leave. Although the rate of improvement has slowed recently the data also suggests an overall downward trend in the wage gap over.
Moreover some countries show fluctuations rather than steady advancement, indicating that policy changes, labour force participation variations all impact wage equity outcomes. Importantly, variations in seniority, industry,hours worked and occupation- factors that can both mask and intensify underlying discrimination are not taken into consideration by the gap indicator.  



<!-- END: Discussion -->

<!-- START: Conclusion -->
##Conclusion

The data confirms that the gender wage gap between 2000 and 2020 is still a major obstacle to economic fairness even though it is gradually closing in many nations. Even though there has been some noticeable progress, particularly in countries where specific policy interventions have been implemented, the pace of development is uneven and frequently stalls in the absence of consistent effort. High-gap countries demonstrate how cultural attitudes, insufficient family policies, and limited access to leadership roles for women continue to perpetuate inequality. By improving social cohesiveness, productivity, and talent utilization, a closing pay gap helps economics as well as individuals. Achieving gender pay equity is thus not just a moral imperative but an economic one.


<!-- END: Conclusion -->


<!-- START: Recommendations -->
##Recommendations

1. Government should enforce pay transparency laws and compel big businesses to submit gender-disaggregated compensation data on a regular basis.

2. Increase access to cost-effective childcare, encourage parental leave uptake, and support flexible work schedules to support working parents.

3. Encourage the presence of women in executive and managerial positions by implementing fair promotion practices, sponsorship, quotas.

4. Keep a close eye on developments in the pay gap and assess the success of measures using internationally recognized metrics.

5. Promote STEM education, mentorship and discrimination-free hiring practices to encourage women and girls to pursue historically male-dominated fields.



<!-- END: Recommendations -->
