---
title: "Gender Pay Gap across countries: Industry-Level Trends and Analysis"
subtitle: "Muhammad Nor Hakim bin Mad Yusof 28693310 / Chanya Kitprapa 34684824 / Janhavi Bayanwar 35579331"
format: revealjs
editor: visual
---

## Problem introduction

-   Investigate how and why gender pay gaps differ across countries

-   Explore factors influencing wage gaps using OECD and World Bank data

---

## Dataset description

-   7 countries: Australia, UK, US, Sweden, Japan, Korea, EU\
-   Timeframe: 2000–2020\
-   Data sources: OECD, World Bank\
-   Key variables:
    -   Gender wage gap (%)
    -   GDP per capita (USD, 2015)
    -   Female labor force participation
    -   Part-time employment (%)
    -   Female education (Bachelor+)

---

## Methods

-   Data from 7 regions (e.g., Australia, US, Japan, Sweden, EU)

-   Timeframe: 2000–2020

-   Variables used:

    -   Gender wage gap (% of male earnings)

    -   GDP per capita (USD, 2015)

    -   Female labour force participation (15–64)

    -   Female educational attainment (Bachelor+)

    -   Part-time female employment rate

-   Tools: R, tidyverse, ggplot2, Quarto

## Results / Key Findings – Trends Over Time

-   Gender wage gaps **declined** in all countries studied

-   **Korea and Japan**: highest, slowest to improve

-   **Sweden and EU**: lowest gaps, most consistent improvements

-   **Australia, UK, US**: moderate progress with fluctuations

```{r}
#| echo: false

library(fpp3)
library(kableExtra)
library(dplyr)
library(stringr)

# Data cleaning and preperation

# Gender wage gap data

gender_wage_gap <- read.csv("Data/gender wage gap across time.csv") %>%
  select(Reference.area, OBS_VALUE, TIME_PERIOD) %>% 
  rename(Country = Reference.area, Gender_Wage_Gap = OBS_VALUE, Year = TIME_PERIOD) %>% 
  arrange(Year)

gender_wage_gap <- gender_wage_gap %>%
  group_by(Country, Year) %>%
  summarise(Gender_Wage_Gap = mean(Gender_Wage_Gap), .groups = "drop")

gender_wage_gap <- gender_wage_gap %>%
  mutate(Country = ifelse(Country == "European Union (27 countries)", "European Union", Country))

# GDP per capita data
gdp <- read.csv("Data/gdp per capita.csv") %>%
  select(-X.1) %>% 
  rename(Country = X)

gdp <- gdp %>%
  pivot_longer(
    cols = starts_with("X"),
    names_to = "Year",
    values_to = "GDP_per_capita"
  ) %>%
  mutate(year = as.integer(sub("X", "", Year))) 

gdp <- gdp %>% 
  select(Country, GDP_per_capita, year) %>% 
  rename(Year = year)

# Merge the two datasets
data <- left_join(gender_wage_gap, gdp, by = c("Country", "Year"))

# Gender stats data

gender <- read.csv("Data/gender stats.csv") 


gender <- gender[-c(15,16,17,18,19), ]

gender <- gender %>% 
  select(-Country.Code, -Series.Code) %>% 
  rename(y = Series.Name, Country = Country.Name)

# Labor force participation, female (% of female pop aged 15-64)
labor_force_participation <- gender %>% 
  filter(y == "Labor force participation rate, female (% of female population ages 15-64) (modeled ILO estimate)") %>% 
  select(-y) %>% 
  pivot_longer(
    cols = -Country,
    names_to = "Year",
    values_to = "Labour_force_participation"
  ) %>%
  mutate(Year = as.integer(sub(".*?(\\d{4}).*", "\\1", Year))) 

# Part time employment, female (% of total female employment)
part_time_employment <- gender %>% 
  filter(y == "Part time employment, female (% of total female employment)") %>% 
  select(-y) %>% 
  pivot_longer(
    cols = -Country,
    names_to = "Year",
    values_to = "Part_time_employment"
  ) %>%
  mutate(Year = as.integer(sub(".*?(\\d{4}).*", "\\1", Year)))
  
# Merge datasets
data <- left_join(data, labor_force_participation, by = c("Country", "Year"))

data <- left_join(data, part_time_employment, by = c("Country", "Year")) 

female_bachelors <- read.csv("Data/female bachelors attainment.csv") %>% 
  rename(Country = X) %>% 
  select(-X.1) %>% 
  mutate(Country = ifelse(Country == "Korea, Rep.", "Korea", Country)) %>% 
  pivot_longer(
    cols = starts_with("X"),
    names_to = "Year",
    values_to = "Female_bachelors") %>%
  mutate(Year = as.integer(sub("X", "", Year))) 

# Merge datasets
data <- left_join(data, female_bachelors, by = c("Country", "Year"))

# Convert data str
data <- data %>% 
  mutate(Labour_force_participation = as.numeric(Labour_force_participation))

```

```{r}
#| label: fig-medianwagegap
#| fig-cap: "Median gender wage gap by country (2000-2020)"
#| echo: false

ggplot(data, aes(x = Year, y = Gender_Wage_Gap, color = Country)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  labs(
    title = "Gender Wage Gap",
    x = "Year",
    y = "Median Gender Wage Gap (%)",
    color = "Country"
  ) +
  theme_minimal()

```

------------------------------------------------------------------------

## GDP vs Wage Gap

-   No clear relationship between **high GDP** and **low wage gaps**

-   Example: US and Australia have high GDP but still notable wage gaps

-   Structural factors matter more than income level

```{r}
#| label: fig-wagegapvsgdp
#| fig-cap: "Median gender wage gap vs GDP per capita across countries (2000-2020)"
#| echo: false

data_filtered <- data %>% 
  filter(Country !="Korea")

data_gdp <- data_filtered %>%
  mutate(GDP_per_capita = str_replace_all(GDP_per_capita, "[^0-9.]", ""),
         GDP_per_capita = as.numeric(GDP_per_capita))

ggplot(data_gdp, aes(x = GDP_per_capita, y = Gender_Wage_Gap, colour = Country)) +
  geom_point(size = 4) +
  labs(
    x = "GDP per Capita (2015, $US)",
    y = "Median Gender Wage Gap (%)",
    color = "Country"
  ) +
  theme_minimal()
```

------------------------------------------------------------------------

## Discussion

-   Wage gaps persist due to **structural inequalities**

-   Cultural norms, family leave policies, and workplace promotion bias

-   Gaps are **not explained by GDP alone**

-   Indicator does not adjust for hours worked, role seniority, or industry

------------------------------------------------------------------------

## Conclusion

-   Gender wage gaps are narrowing overall, but **unevenly**

-   Progress depends on consistent, structural policy interventions

-   **Economic equity** benefits individuals and the broader economy

------------------------------------------------------------------------

## Recommendations

-   Enforce pay transparency reporting

-   Support childcare access and flexible work

-   Promote women in leadership roles

-   Monitor wage gap metrics with global standards

-   Encourage STEM education and fair hiring practices
